---
title: "Interactive Report"
format: html
runtime: shiny
---

```{r setup, include=FALSE}
library(shiny)
library(ggplot2)
library(datasets)
library(dplyr)
```

## Descriptive Statistics

This section includes static descriptive charts for the `mtcars` dataset.

### Distribution of Miles per Gallon (MPG)

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "blue", color = "white") +
  labs(title = "Histogram of Miles per Gallon (MPG)", x = "MPG", y = "Count")
```

### Relationship between MPG and Weight

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "darkred", size = 2) +
  labs(title = "Scatter Plot of MPG vs Weight", x = "Weight (1000 lbs)", y = "MPG")
```

## Interactive Component: K-Means Clustering

Use the slider below to adjust the number of clusters in the k-means algorithm. The plot will update dynamically based on your selection.

### Slider Input for Clusters

```{r}
sliderInput("clusters", "Number of Clusters:", min = 2, max = 10, value = 3)
```

### Reactive Plot

```{r}
renderPlot({
  set.seed(123)
  num_clusters <- input$clusters
  kmeans_result <- kmeans(mtcars[, c("mpg", "wt")], centers = num_clusters)
  mtcars$cluster <- factor(kmeans_result$cluster)

  ggplot(mtcars, aes(x = mpg, y = wt, color = cluster)) +
    geom_point(size = 3) +
    labs(
      title = paste("K-Means Clustering with", num_clusters, "Clusters"),
      x = "Miles per Gallon (MPG)",
      y = "Weight (1000 lbs)"
    ) +
    theme_minimal()
})
```

### Output Plot

```{r}
plotOutput("kmeans_plot")
```
